---
title: "Binomial Proportion Confidence Interval"
author: "Art Eschenlauer"
date: "January 23, 2019"
output:
  pdf_document: 
    toc: yes
---


## Introductory remarks

Many ways have been developed to estimate the confidence interval for 
an observed propoortion.
See 
[https://en.wikipedia.org/wiki/Binomial_proportion_confidence_interval](https://en.wikipedia.org/wiki/Binomial_proportion_confidence_interval)
for a list of them.
Agresti and Coull (1998) recommend that elementary statistics instruction
prefer the "Score" method of Wilson (1927) for estimating the confidence
interval over the simpler, but insufficiently conservative, normal
approximation.  I prefer the Wilson Score method because it performs well
for proportions that approach 0 or 1, where the normal approximation 
breaks down.  It is a bit more complicated to calculate.


## A sanity check

Before computing the confidence interval, I first will take some wild
guesses at the probabilities near the boundaries.  For my example, I will
assume that 73 successes were observed for 76 trials, and that I want a
95% confidence interval.

I manually adjusted my guesses to obtain a 2.5% area under probability 
density function (PDF) curve for each tail.  This would require 
integration, but the cumulative density function (CDF) is the integration
of the PDF.
For R, the `stats::pbinom` function implements CDF for the binomial distribution.

So, what value do I expecting to be near the lower limit of the CI?

```{r}
successes <- 73
trials <- 76
proportion <- as.double(successes) / as.double(trials)
pbinom(q = successes, size = trials, prob = proportion)
low.guess <- 0.90816
high.guess <- 0.99178
pbinom(q = successes, size = trials, prob = low.guess)
```

And, what value do I expecting to be near the upper limit of the CI?

```{r}
pbinom(q = successes, size = trials, prob = high.guess, lower.tail = FALSE)
```

## The Wilson Score Interval - from Wilson 1927

The 1927 paper by Edwin Bidwell Wilson presented the following formula as a 
practical way to estimate the confidence interval for a proportion:

$$
\newcommand{\relmiddle}[1]{\mathrel{}\middle#1\mathrel{}}
{
\left. 
p = \frac{p_0 + \frac{t}{2} \pm \sqrt{p_0q_0t + \frac{t^2}{4}}}{1 + t} 
\relmiddle| 
\begin{aligned}\
\lambda &= 1.96 \text{ (for 95 percent confidence)},\\
t &= \frac{\lambda^2}{n}, \quad
p_0 = \frac{x}{n}, \quad
q_0 = \frac{n - x}{n}\\
\end{aligned}
\right.
}
$$

The CRAN R package `epitools` provides the `binom.wilson` function which can compute this:

```{r}
epi_wilson_ci <- 
  epitools::binom.wilson(x = successes, n = trials, conf.level = 0.95)
print(epi_wilson_ci)
```

To demystify the `epitools` implemntation, I coded it directly from the equation above: 

```{r}
# computational equivalent of binom.wilson from CRAN 'epitools' package
binom.wilson <-
  function(x, n, conf.level = 0.95) {
    lambda  <- qnorm(0.5*(1 + conf.level))
    # Adapted from Wilson 1927, mostly by multipling numerator
    #   and denominator by n to reduce roundoff error
    n_t     <- ( lambda^2 )
    n_p_0   <- x
    n_q_0   <- n - x
    denom   <- n + n_t
    center  <- ( n_p_0 + n_t / 2 )
    radical <- ( n_p_0 * n_q_0 + n * (n_t / 4) ) * n_t / n
    # continuing
    delta   <- sqrt(radical)
    R.lower <- ( center - delta ) / denom
    R.upper <- ( center + delta ) / denom
    data.frame(
      x = x,
      n = n,
      proportion = as.double(n_p_0) / as.double(n), 
      lower = R.lower,
      upper = R.upper,
      conf.level = conf.level
    )
  }
wilson_ci <- binom.wilson(x = successes, n = trials, conf.level = 0.95)
print(wilson_ci)
```

There is negligible difference between the `epitools` implementation and my own: 

```{r}
print(wilson_ci$lower - epi_wilson_ci$lower)
print(wilson_ci$upper - epi_wilson_ci$upper)
```

I have not contrasted the two implementations to see which suffers more from round-off error, but it seems clear that they are consistent with one another.

## Conclusions 

```{r}
print(
  sprintf("For %d successes over %d trials:"
  , wilson_ci$x
  , wilson_ci$n
  )
)
print(
  sprintf("Proportion: %f"
  , wilson_ci$proportion
  )
)
print(
  sprintf("Wilson estimate of confidence interval for proportion: [%f,%f]"
  , wilson_ci$lower
  , wilson_ci$upper
  )
)
print(
  sprintf("My low and high guesses: %f,%f"
  , low.guess
  , high.guess
  )
)
print(
  sprintf("Differences from my guesses: %f, %f"
  , wilson_ci$lower - low.guess
  , wilson_ci$upper - high.guess
  )
)
```

So, this seems to me to pass the sanity checks.  I think that the results are not the same at least in part because the sanity checks are based on different actual proportions.

## About this document

This document was produced from an RMarkdown file, the source of which should be at:

[https://eschen42.github.io/propci/propci.Rmd](https://eschen42.github.io/propci/propci.Rmd)

## References

Agresti, Alan; Coull, Brent A. (1998). "Approximate is better than 'exact' for interval estimation of binomial proportions". The American Statistician. 52: 119-126. [doi:10.2307/2685469](https://doi.org/10.2307/2685469).

Wilson, E. B. (1927). "Probable inference, the law of succession, and statistical inference". Journal of the American Statistical Association. 22: 209-212. [doi:10.1080/01621459.1927.10502953](https://doi.org/10.1080/01621459.1927.10502953).